---
title: "Longitudinal Analysis for final project"
author: "Shih-Ni Prim"
date: "11/12/2020"
fontsize: 12pt
geometry: margin = 1in
mainfont: Times New Roman
spacing: 1.5
fig_caption: yes
indent: true
header-includes:
  - \usepackage{endfloat}  
output: 
  bookdown::pdf_document2:
    latex_engine: lualatex
    toc: true
    fig_width: 6.5
    fig_height: 4
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE)
library(readxl)
library(tidyverse)
library(ggplot2)
library(httr)
library(nlme)
library(emmeans)
library(GGally)
library(knitr)

library(bookdown)
```

```{r readin Data, results = 'hide'}
#Readin data from Github url address
url1 <- 'https://github.com/luokan1227/537P1/raw/master/Data.xlsx'  # main data
url2 <- 'https://github.com/luokan1227/537P1/raw/master/MonkeyID.xlsx'  # Monkey ID
GET(url1, write_disk(tf1 <- tempfile(fileext = ".xlsx")))
GET(url2, write_disk(tf2 <- tempfile(fileext = ".xlsx")))
Rawdata <- read_excel(tf1)
MonkeyID <-  read_excel(tf2)
#Add Monkey ID to raw data
Rawdata <- inner_join(Rawdata, MonkeyID, by = "H_ID")
colnames(Rawdata)[2] <- "Time_Point"
#-----------------------
#Adjust or add variables
#-----------------------

# Final data set for all following analysis
Data <- Rawdata

# add drug type and reactivity
Data$Drug <- ifelse(Data$Treatment == "group 7", 3, 
                      ifelse(Data$Treatment == "group 1", 1, 
                      ifelse(Data$Treatment == "group 2", 1, 
                      ifelse(Data$Treatment == "group 3", 1, 2))))

Data$Reactivity <- ifelse(Data$Binding > 0.1, 1, 0)

# create a subset with the variables we need plus the extracted information
Data2 <- Data %>% select(MonkeyID, Drug, Treatment, Time_Point, Isotype, H_VBase, H_Substitutions, H_Insertions, H_Deletions, HMuFreq, H_CDR3, L_VBase, L_Substitutions, L_Insertions, L_Deletions, LMuFreq, L_CDR3, Binding, Reactivity)

# final data 
Data3 <- Data2[-972,]
```

3. **How does the binding strength of the antibodies develop in response to the number of vaccine dosages by treatment?** This will be evaluated with a longitudinal analysis by test subject.  

# Data Analysis  

## Longitudinal Data Analysis  

As seen in Figure \@ref(fig:mean-monkey) and Figure \@ref(fig:variance-monkey), the mean trend is not linear, and the different time points have different variances. This information suggests that we should use piecewise linear models and set variances as unequal over time.  


```{r, mean-monkey, fig.cap = "Mean trend by monkey"}
ggplot(Data3, aes(x = Time_Point, y = Binding, colour = as.factor(MonkeyID))) + geom_point() + stat_summary(fun = mean, geom = "line")
```

```{r, variance-monkey, fig.cap = "Variances over time by monkey"}
ggplot(Data3, aes(x = Time_Point, y = Binding, colour = as.factor(MonkeyID))) + geom_point() + stat_summary(fun = var, geom = "line")
```


We first consider a model with time point as the only covariate: $$Y_{ij}=\beta_0+\beta_1 Time_{ij}+e_{ij}$$  


Thus we will use a piecewise linear model, in which each segment has different intercepts and slopes. We use three indicator variables: $S1, S2, S3$ as the indicator variables, where 
$$S1=
\begin{cases}
1 & \text{if } 0 \le \text{Timepoint} < 1 \\
0 & \text{otherwise}
\end{cases}
$$
$$S2=
\begin{cases}
1 & \text{if } 1 \le \text{Timepoint} < 2 \\
0 & \text{otherwise}
\end{cases}
$$
$$S3=
\begin{cases}
1 & \text{if Timepoint} \ge 2 \\
0 & \text{otherwise}
\end{cases}
$$

The new model is thus $$Y_{ij}=S1(\beta_0+\beta_1 Time_{ij})+S2(\beta_2+\beta_3 Time_{ij})+S3(\beta_4+\beta_5 Time_{ij})+e_{ij}$$  

We also want to make sure that the trend is continuous at timepoint = 1 and 2.  

Our final model is $Y_{ij} = \beta_0(S1 + 2S2 - S2 Time_{ij}) + \beta_1 (S1 Time_{ij} + 2S2 - S2 Time_{ij}) + \beta_4(-S2 + S2 Time_{ij} + S3) + \beta_5(-2S2 + 2S2 Time_{ij} + S3 Time_{ij}) + e_{ij}$ where $$\mathbf{e}_i \sim N(0, \sigma^2I)$$  

```{r}
S1 <- as.numeric(Data3$Time_Point >= 0 & Data3$Time_Point < 1)
S2 <- as.numeric(Data3$Time_Point >= 1 & Data3$Time_Point < 2)
S3 <- as.numeric(Data3$Time_Point >= 2)
dataLDA1 <- data.frame(id = Data3$MonkeyID, binding = Data3$Binding, time = Data3$Time_Point, S1 = S1, S2 = S2, S3 = S3)
meanform <- binding ~ -1 + I(S1 + 2*S2 - S2:time) + I(S1:time + 2*S2 - S2:time) + I(-S2 + S2:time + S3) + I(-2*S2 + 2*S2:time + S3:time)
fit.gls <- gls(model = meanform, data = dataLDA1, weights = varIdent(form = ~ 1 | time), correlation = corCompSymm(form = ~1 | id))
summary(fit.gls)
```


S1: $-0.2221651+0.2432183*time$  
S2: $(2*-0.2221651+2*0.2432183-0.7699600+2*0.2432756)+(0.2221651-0.2432183+0.7699600-2*0.2432756)*time = -0.2413024 + 0.2623556*time$  
S3: $0.7699600-0.2432756*time$  

```{r}
# when S1 = 1
f1 <- function(x, y){y = -0.2221651+0.2432183*x}
# when S2 = 1
f2 <- function(x, y){y = -0.2413024 + 0.2623556*x}
# when S3 = 1
f3 <- function(x, y){y = 0.7699600-0.2432756*x}
```
```{r, piecewise1, fig.cap = "Piecewise Linear Function"}
ggplot(Data3, aes(x = Time_Point, y = Binding, colour = as.factor(MonkeyID))) + 
  geom_point() + stat_function(fun = f1, xlim = c(0, 1), geom = "line") + 
  geom_point() + stat_function(fun = f2, xlim = c(1, 2), geom = "line") + 
  geom_point() + stat_function(fun = f3, xlim = c(2, 3), geom = "line")
```

Next we check whether adding random effects improve the model. We assume that random effects exist in the intercept and slope. Our linear mixed model is then: $Y_{ij} = \beta_0(S1 + 2S2 - S2 Time_{ij}) + \beta_1 (S1 Time_{ij} + 2S2 - S2 Time_{ij}) + \beta_4(-S2 + S2 Time_{ij} + S3) + \beta_5(-2S2 + 2S2 Time_{ij} + S3 Time_{ij}) + b_{0i} + b{1i} Time_{ij} + e_{ij}$  
where $$\mathbf{b}_i \sim N \left( 0, \mathbf{D} = 
\begin{pmatrix}
D_{11} & D_{12} \\
 & D_{22}
 \end{pmatrix} \right)$$ and $$\mathbf{e}_i \sim N(0, \sigma^2I)$$  

```{r}
# with random intercept and slope, compound symmetry, unequal variances
fit.a1 <- lme(fixed = meanform, random = ~ time | id, data = dataLDA1, weights = varIdent(form = ~ 1 | time), correlation = corCompSymm(form = ~1 | id))
summary(fit.a1)
```

```{r}
# random intercept and slope, AR1, unequal variances
fit.a2 <- lme(fixed = meanform, random = ~ time | id, data = dataLDA1, weights = varIdent(form = ~ 1 | time), correlation = corAR1(form = ~ 1 | id))
```

```{r}
AIC(fit.gls, fit.a1, fit.a2)
BIC(fit.gls, fit.a1, fit.a2)
```

We now check residuals for both models.  

```{r}
par(mfrow = c(1,3))
gls.res <- residuals(fit.gls, type = "normalized")
qqnorm(gls.res)
lme.res1 <- residuals(fit.a1, level = 1, type = "normalized")
qqnorm(lme.res1)
lme.res2 <- residuals(fit.a2, level = 1, type = "normalized")
qqnorm(lme.res2)
```



## With Drugs  

```{r, LDA-overview, fig.cap = "Means and Variances over timepoints"}
# simply connects the mean of each time point
ggplot(Data3, aes(x = Time_Point, y = Binding)) + geom_point(stat = "identity", aes(color = as.factor(Drug))) + geom_jitter() + stat_summary(fun = mean, geom = "line", lwd = 2, color = "red") + stat_summary(fun = var, geom = "line", lwd = 2, color = "blue") + scale_color_discrete(name = "Drug")

```

Here we use `Binding` as the response, `Time_Point` as the time factor, and `Drug` as the covariates. Random effect for both intercept and slope. Now we want to add one covariate: `Drug`. We use two indicator variables: `D1` and `D2`, where  

$$D1=
\begin{cases}
1 & \text{if Drug = 1} \\
0 & \text{otherwise}
\end{cases}
$$

$$D2=
\begin{cases}
1 & \text{if Drug = 2} \\
0 & \text{otherwise}
\end{cases}
$$

Assuming that the random effects are the same for each drug, our full model is: 

$$Y_{ij} = \beta_0+ \beta_1 Time_{ij} + D1_i(\beta_2+\beta_3 Time_{ij}) + D2_i (\beta_4+\beta_5 Time_{ij}) + b_{0i} + b_{1i} Time_{ij} + e_{ij}$$  
$$\mathbf{b_i} \sim N \left(0, \mathbf{D}=
\begin{bmatrix}
D_{11} & D_{12} \\
D_{12} & D_{22} 
\end{bmatrix}
\right)$$

Drug 1: $Y_{ij} = \beta_0+\beta_1 Time_{ij} + \beta_2+\beta_3 Time_{ij} + b_{0i} + b_{1i} Time_{ij} + e_{ij}$  
Drug 2: $Y_{ij} = \beta_0+\beta_1 Time_{ij} + \beta_4+\beta_5 Time_{ij} + b_{0i} + b_{1i} Time_{ij} + e_{ij}$  
Drug 3: $Y_{ij} = \beta_0+\beta_1 Time_{ij} + b_{0i} + b_{1i} Time_{ij} + e_{ij}$  

```{r}
dataLDA <- data.frame(id = Data3$MonkeyID, binding = Data3$Binding, Time_Point = Data3$Time_Point, D1 = as.numeric(Data3$Drug == 1), D2 = as.numeric(Data3$Drug == 2))
# head(dataLDA)

lda <- lme(fixed = binding ~ Time_Point + D1 + D1:Time_Point + D2 + D2:Time_Point,
           random = ~ Time_Point | id, weights = varIdent(form = ~ 1 | Time_Point), data = dataLDA, method = "REML")
summary(lda)
```

The p-values for `Drug` and the interaction of `Drug` and `Time_Point` are large. So we try another model with `Time_Point` as the only predictor.  [This is skipping the part where we fit only main effect (not interaction) with Drug]

$$Y_{ij} = \beta_0+ \beta_1 Time_{ij} + b_{0i} + b_{1i} Time_{ij} + e_{ij}$$  

$$
\underbrace{\begin{bmatrix}
Y_{i1} \\
\vdots \\
Y_{im_i}
\end{bmatrix}}_{\mathbf{Y_i}}
=
\underbrace{\begin{bmatrix}
1 & Time_{i1} \\
\vdots & \vdots & \vdots \\
1 & Time_{im_i} & 
\end{bmatrix}}_{\mathbf{X_i}}
\underbrace{\begin{bmatrix}
\beta_0 \\
\beta_1 
\end{bmatrix}}_{\mathbf{\beta}}
+
\underbrace{\begin{bmatrix}
1 & Time_{i1} \\
\vdots & \vdots \\
1 & Time_{im_i}
\end{bmatrix}}_{\mathbf{Z_i}} 
\underbrace{\begin{bmatrix}
b_{0i} \\
b_{1i}
\end{bmatrix}}_{\mathbf{b_i}}
+ 
\underbrace{\begin{bmatrix}
e_{i1} \\
\vdots \\
e_{im_i}
\end{bmatrix}}_{\mathbf{e_i}}$$

$$\mathbf{b_i} \sim N \left(0, \mathbf{D}=
\begin{bmatrix}
D_{11} & D_{12} \\
D_{12} & D_{22} 
\end{bmatrix}
\right)$$  

$$\mathbf{e}_{ij} \sim N(0, \mathbf{R}_i = \sigma^2 I_{mi})$$

[need to consider whether time point 2 is the optimal point]

```{r}
lda2 <- lme(fixed = binding ~ Time_Point,
           random = ~ Time_Point | id, weights = varIdent(form = ~ 1 | Time_Point), data = dataLDA, method = "REML")
summary(lda2)
```

This simpler model has lower AIC and BIC, as shown below. So we prefer the model with `Time_Point` as the predictor and, with the low p-values of the slope of `Time_Point`, conclude that the binding rates vary over time. In other words, the number of HIV vaccines given do affect the binding rate, but the drugs given do not have significant effects. As seen in Table \@ref(tab:AIC-table), blah blah...  


```{r, AIC-table}
knitr::kable(data.frame(cbind(AIC(lda, lda2), BIC(lda, lda2))), caption = "AIC and BIC for Longitudinal Models")
```

