---
title: "Longitudinal Analysis for final project"
author: "Shih-Ni Prim"
date: "11/12/2020"
fontsize: 12pt
geometry: margin = 1in
mainfont: Times New Roman
spacing: 1.5
fig_caption: yes
indent: true
header-includes:
  - \usepackage{endfloat}  
output: 
  bookdown::pdf_document2:
    latex_engine: lualatex
    toc: true
    fig_width: 6.5
    fig_height: 4
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE)
library(readxl)
library(tidyverse)
library(ggplot2)
library(httr)
library(nlme)
library(emmeans)
library(GGally)
library(knitr)
library(clubSandwich)
library(bookdown)
```

```{r readin Data, results = 'hide'}
#Readin data from Github url address
url1 <- 'https://github.com/luokan1227/537P1/raw/master/Data.xlsx'  # main data
url2 <- 'https://github.com/luokan1227/537P1/raw/master/MonkeyID.xlsx'  # Monkey ID
GET(url1, write_disk(tf1 <- tempfile(fileext = ".xlsx")))
GET(url2, write_disk(tf2 <- tempfile(fileext = ".xlsx")))
Rawdata <- read_excel(tf1)
MonkeyID <-  read_excel(tf2)
#Add Monkey ID to raw data
Rawdata <- inner_join(Rawdata, MonkeyID, by = "H_ID")
colnames(Rawdata)[2] <- "Time_Point"
#-----------------------
#Adjust or add variables
#-----------------------

# Final data set for all following analysis
Data <- Rawdata

# add drug type and reactivity
Data$Drug <- ifelse(Data$Treatment == "group 7", 3, 
                      ifelse(Data$Treatment == "group 1", 1, 
                      ifelse(Data$Treatment == "group 2", 1, 
                      ifelse(Data$Treatment == "group 3", 1, 2))))

Data$Reactivity <- ifelse(Data$Binding > 0.1, 1, 0)

# create a subset with the variables we need plus the extracted information
Data2 <- Data %>% select(MonkeyID, Drug, Treatment, Time_Point, Isotype, H_VBase, H_Substitutions, H_Insertions, H_Deletions, HMuFreq, H_CDR3, L_VBase, L_Substitutions, L_Insertions, L_Deletions, LMuFreq, L_CDR3, Binding, Reactivity)

# final data 
Data3 <- Data2[-972,]
```

```{r}
Data4 <- Data3 %>% select(H_CDR3, HMuFreq, L_CDR3, LMuFreq, Binding, Drug, Isotype, Time_Point)
Data4_drug1 <- Data4 %>% filter(Drug == 1)
cov1 <- Data4_drug1 %>% select(H_CDR3, HMuFreq, L_CDR3, LMuFreq, Binding) %>% cov()
Data4_drug2 <- Data4 %>% filter(Drug == 2)
cov2 <- Data4_drug2 %>% select(H_CDR3, HMuFreq, L_CDR3, LMuFreq, Binding) %>% cov()
Data4_drug3 <- Data4 %>% filter(Drug == 3)
cov3 <- Data4_drug3 %>% select(H_CDR3, HMuFreq, L_CDR3, LMuFreq, Binding) %>% cov()
cov1 / cov2
cov1 / cov3
cov2 / cov3
Data4_isog <- Data4 %>% filter(Isotype == "G")
covg <- Data4_isog %>% select(H_CDR3, HMuFreq, L_CDR3, LMuFreq, Binding) %>% cov()
Data4_isom <- Data4 %>% filter(Isotype == "M")
covm <- Data4_isom %>% select(H_CDR3, HMuFreq, L_CDR3, LMuFreq, Binding) %>% cov()
covg / covm

Data5 <- Data3 %>% select(Drug, Isotype)
Data5$Binding <- scale(Data3$Binding)
Data5$H_CDR3 <- scale(Data3$H_CDR3)
Data5$HMuFreq <- scale(Data3$HMuFreq)
Data5$L_CDR3 <- scale(Data3$L_CDR3)
Data5$LMuFreq <- scale(Data3$LMuFreq)
bartlett.test(Binding ~ Drug, data = Data5)
bartlett.test(H_CDR3 ~ Drug, data = Data5)
bartlett.test(HMuFreq ~ Drug, data = Data5)
# only these two are not significant 
bartlett.test(L_CDR3 ~ Drug, data = Data5)
bartlett.test(LMuFreq ~ Drug, data = Data5)
##
bartlett.test(Binding ~ Isotype, data = Data5)
bartlett.test(H_CDR3 ~ Isotype, data = Data5)
bartlett.test(HMuFreq ~ Isotype, data = Data5)
bartlett.test(L_CDR3 ~ Isotype, data = Data5)
bartlett.test(LMuFreq ~ Isotype, data = Data5)
bartlett.test(Binding ~ Data3$Time_Point, data = Data5)
bartlett.test(H_CDR3 ~ Data3$Time_Point, data = Data5)
bartlett.test(HMuFreq ~ Data3$Time_Point, data = Data5)
bartlett.test(L_CDR3 ~ Data3$Time_Point, data = Data5)
bartlett.test(LMuFreq ~ Data3$Time_Point, data = Data5)


tp <- as.factor(Data3$Time_Point)
it <- as.factor(Data3$Isotype)
drug <- as.factor(Data3$Drug)
fit.manova1 <- manova(cbind(Data3$L_CDR3, Data3$LMuFreq) ~ drug)
summary(fit.manova1, test = "Wilks")
```


# Data Analysis  

## Longitudinal Data Analysis  

To answer our research question Q2, **How does the binding strength of the antibodies develop in response to the number of vaccine dosages by treatment?**, we use longitudinal data analysis, including general linear models and linear mixed models.

As seen in Figure \@ref(fig:mean-monkey) and Figure \@ref(fig:variance-monkey), the mean trend is not linear, and the different time points have different variances. This information suggests that we should use piecewise linear models and set variances as unequal over time.  


```{r, mean-monkey, fig.cap = "Mean trend by monkey"}
ggplot(Data3, aes(x = Time_Point, y = Binding, colour = as.factor(MonkeyID))) + geom_point() + stat_summary(fun = mean, geom = "line")
```

```{r, variance-monkey, fig.cap = "Variances over time by monkey"}
ggplot(Data3, aes(x = Time_Point, y = Binding, colour = as.factor(MonkeyID))) + geom_point() + stat_summary(fun = var, geom = "line")
```


We first consider a model with time point as the only covariate: $$Y_{ij}=\beta_0+\beta_1 Time_{ij}+e_{ij}$$  


Thus we will use a piecewise linear model, in which each segment has different intercepts and slopes. We use three indicator variables: $S1, S2, S3$ as the indicator variables, where 
$$S1=
\begin{cases}
1 & \text{if } 0 \le \text{Timepoint} < 1 \\
0 & \text{otherwise}
\end{cases}
$$
$$S2=
\begin{cases}
1 & \text{if } 1 \le \text{Timepoint} < 2 \\
0 & \text{otherwise}
\end{cases}
$$
$$S3=
\begin{cases}
1 & \text{if Timepoint} \ge 2 \\
0 & \text{otherwise}
\end{cases}
$$

The new model is thus $$Y_{ij}=S1(\beta_0+\beta_1 Time_{ij})+S2(\beta_2+\beta_3 Time_{ij})+S3(\beta_4+\beta_5 Time_{ij})+e_{ij}$$  

We also want to make sure that the trend is continuous at timepoint = 1 and 2.  

Our model is $Y_{ij} = \beta_0(S1 + 2S2 - S2 Time_{ij}) + \beta_1 (S1 Time_{ij} + 2S2 - S2 Time_{ij}) + \beta_4(-S2 + S2 Time_{ij} + S3) + \beta_5(-2S2 + 2S2 Time_{ij} + S3 Time_{ij}) + e_{ij}$ where $$\mathbf{e}_i \sim N(0, \sigma^2I)$$  

```{r}
S1 <- as.numeric(Data3$Time_Point >= 0 & Data3$Time_Point < 1)
S2 <- as.numeric(Data3$Time_Point >= 1 & Data3$Time_Point < 2)
S3 <- as.numeric(Data3$Time_Point >= 2)
dataLDA1 <- data.frame(id = Data3$MonkeyID, binding = Data3$Binding, time = Data3$Time_Point, S1 = S1, S2 = S2, S3 = S3)
meanform <- binding ~ -1 + I(S1 + 2*S2 - S2:time) + I(S1:time + 2*S2 - S2:time) + I(-S2 + S2:time + S3) + I(-2*S2 + 2*S2:time + S3:time)
fit.gls <- gls(model = meanform, data = dataLDA1, weights = varIdent(form = ~ 1 | time), correlation = corCompSymm(form = ~1 | id))
# summary(fit.gls)
```

Again, our final mean model is 
$$Y_{ij} = \beta_0(S1 + 2S2 - S2 Time_{ij}) + \beta_1 (S1 Time_{ij} + 2S2 - S2 Time_{ij})+$$
$$\beta_4(-S2 + S2 Time_{ij} + S3) + \beta_5(-2S2 + 2S2 Time_{ij} + S3 Time_{ij}) + e_{ij}$$ which can be written as $$Y_{ij} = S1(\beta_0) + S1 Time_{ij} (\beta_1) + S2( 2\beta_0 + 2\beta_1 - \beta_4 -2\beta_5) + S2 Time_{ij}(-\beta_0 -\beta_1 + \beta_4 + 2\beta_5)$$
$$ + S3(\beta_4) + S3 Time_{ij} (\beta_5) + e_{ij}$$  
From the model above, we can find the intercepts and slopes for all three segments of the mean trend and make a plot, as seen in Figure \@ref(fig:piecewise1):  

S1: $-0.2221651+0.2432183*time$  
S2: $(2*-0.2221651+2*0.2432183-0.7699600+2*0.2432756)+(0.2221651-0.2432183+0.7699600-2*0.2432756)*time = -0.2413024 + 0.2623556*time$  
S3: $0.7699600-0.2432756*time$  

```{r}
# when S1 = 1
f1 <- function(x, y){y = -0.2221651+0.2432183*x}
# when S2 = 1
f2 <- function(x, y){y = -0.2413024 + 0.2623556*x}
# when S3 = 1
f3 <- function(x, y){y = 0.7699600-0.2432756*x}
```

```{r, piecewise1, fig.cap = "Piecewise Linear Function--three segments"}
ggplot(Data3, aes(x = Time_Point, y = Binding, colour = as.factor(MonkeyID))) + 
  geom_point() + stat_function(fun = f1, xlim = c(0, 1), geom = "line") + 
  stat_function(fun = f2, xlim = c(1, 2), geom = "line") + 
  stat_function(fun = f3, xlim = c(2, 3), geom = "line")
```

As shown in Figure \@ref(fig:piecewise1), the two segments S1 and S2 look linear. So now we'll refit the model with only two piecewise sections; we'll call them S4 and S5. The new model is therefore $$Y_{ij}=S4(\beta_0+\beta_1 Time_{ij})+S5(\beta_2+\beta_3 Time_{ij})+e_{ij}$$  

$$S4=
\begin{cases}
1 & \text{if Timepoint} < 2 \\
0 & \text{otherwise}
\end{cases}
$$
$$S5=
\begin{cases}
1 & \text{if Timepoint} \ge 2 \\
0 & \text{otherwise}
\end{cases}
$$

We also want to make sure that the trend is continuous at Time_Point = 2.  

Our model is then $Y_{ij} = \beta_1(-2S4 + S4 Time_{ij}) + \beta_2(S4 + S5) + \beta_3(2S4 + S5 Time_{ij}) + e_{ij}$ where $$\mathbf{e}_i \sim N(0, \sigma^2I)$$  

```{r}
S4 <- as.numeric(Data3$Time_Point < 2)
S5 <- as.numeric(Data3$Time_Point >= 2)
dataLDA1$S4 <- S4
dataLDA1$S5 <- S5
meanform2 <- binding ~ -1 + I(-2*S4 + S4:time) + I(S4 + S5) + I(2*S4 + S5:time)
fit.gls2 <- gls(model = meanform2, data = dataLDA1, weights = varIdent(form = ~ 1 | time), correlation = corCompSymm(form = ~1 | id))
# summary(fit.gls2)
```

Again, our model is $Y_{ij}= \beta_1(-2S4 + S4 Time_{ij}) + \beta_2(S4 + S5) + \beta_3(2S4 + S5 Time_{ij}) + e_{ij}$, which can be written as $Y_{ij} = S4(-2 \beta_1 + \beta_2 + 2\beta_3) + S4 Time_{ij}(\beta_1) + S5(\beta_2) + S5 Time_{ij}(\beta_3) + e_{ij}$  

We first find the mean trend for S4 and S5:  

S4: $(-2*0.5310975 + 0.5720853 + 2*-0.0000723) + 0.5310975*time = -0.4902543 + 0.5310975*time$
S5: $0.5720853 - 0.0000723*time$

We can make the plot again to see if the model is reasonable, as shown in Figure \@ref(fig:piecewise2). Indeed, there is a linear line between Time_Point 0 and 2 and one between Time_Point 2 and 3. The two lines are contiuous at Time_Point 2. A comparison of AIC And BIC of these two models, shown in Table \@ref(tab:AIC-tab1), indicates that the second model (`fit.gls2`) is indeed a better model. We'll add random effects to it.  

```{r}
# when S4 = 1
f4 <- function(x, y){y = -0.4902543 + 0.5310975*x}
# when S5 = 1
f5 <- function(x, y){y = 0.5720853 - 0.0000723*x}
```

```{r, piecewise2, fig.cap = "Piecewise Linear Function--two segments"}
ggplot(Data3, aes(x = Time_Point, y = Binding, colour = as.factor(MonkeyID))) + 
  geom_point() + stat_function(fun = f4, xlim = c(0, 2), geom = "line") + 
  stat_function(fun = f5, xlim = c(2, 3), geom = "line")
```

```{r, AIC-tab1}
knitr::kable(data.frame(cbind(AIC(fit.gls, fit.gls2), BIC(fit.gls, fit.gls2))), caption = "AIC and BIC between two gls models")
```


Next we check whether adding random effects improve the model. We assume that random effects exist in the intercept and slope. Our linear mixed model is then: $Y_{ij}= \beta_1(-2S4 + S4 Time_{ij}) + \beta_2(S4 + S5) + \beta_3(2S4 + S5 Time_{ij}) + b_{0i} + b{1i} Time_{ij} + e_{ij}$  
where $$\mathbf{b}_i \sim N \left( 0, \mathbf{D} = 
\begin{pmatrix}
D_{11} & D_{12} \\
 & D_{22}
 \end{pmatrix} \right)$$ and $$\mathbf{e}_i \sim N(0, \sigma^2I)$$  

```{r}
# with random intercept and slope, compound symmetry, unequal variances
fit.a1 <- lme(fixed = meanform2, random = ~ time | id, data = dataLDA1, weights = varIdent(form = ~ 1 | time), correlation = corCompSymm(form = ~1 | id))
# summary(fit.a1)
```

```{r}
# random intercept and slope, AR1, unequal variances
fit.a2 <- lme(fixed = meanform2, random = ~ time | id, data = dataLDA1, weights = varIdent(form = ~ 1 | time), correlation = corAR1(form = ~ 1 | id))
```

```{r, AIC-tab2}
kable(data.frame(AIC(fit.gls2, fit.a1, fit.a2), BIC(fit.gls2, fit.a1, fit.a2)), caption = "AIC and BIC for three models")
```

As shown in Table \@ref(tab:AIC-tab2), the model `fit.a2` (random intercept and slope, AR1 correlation, unequal variances) has the lowest AIC And BIC, so it seems the best model. We now check residuals for both models.  

```{r, QQmodel, fig.cap = "Q-Q plots of models: GLS, compound symmetry, AR1"}
par(mfrow = c(1,3))
gls.res <- residuals(fit.gls2, type = "normalized")
qqnorm(gls.res)
lme.res1 <- residuals(fit.a1, level = 1, type = "normalized")
qqnorm(lme.res1)
lme.res2 <- residuals(fit.a2, level = 1, type = "normalized")
qqnorm(lme.res2)
```

All of the Q-Q plots in Figure \@ref(fig:QQmodel) are reasonable, so we'll use `fit.a2` for further analysis.  

Now we would like to know if the slopes between Time_Point 0 and 2 and between Time_Point 2 and 3 equal zero. $H_0: \text{ slope of } S4=0 \text{ and slope of } S5=0$, which means $H_0: \beta_1 = 0 \text{ and } \beta_3=0$  

Thus, we can check for two tests: $$\mathbf{L_1 \beta}=0$$ where $\mathbf{L_1}=(1, 0, 0)$ and $\mathbf{\beta}=(\beta_1, \beta_2, \beta_3)^T$ and $$\mathbf{L_2 \beta} = 0$$ where $\mathbf{L_2}=(0, 0, 1)$ and $\mathbf{\beta}=(\beta_1, \beta_2, \beta_3)^T$  

```{r, S4S5-slope}
L1 <- matrix(c(1, 0, 0), byrow = TRUE, nrow = 1)
L2 <- matrix(c(0, 0, 1), byrow = TRUE, nrow = 1)
test1 <- anova.lme(fit.a2, L = L1, adjustSigma = TRUE)
test2 <- anova.lme(fit.a2, L = L2, adjustSigma = TRUE)
kable(data.frame(rbind(test1, test2)), caption = "Inference of S4 ad S5 slopes")
```

As shown in Table \@ref(tab:S4S5-slope), the slop of S4 has a very small p-value, while the slope of S5 is quite large, indicating that the change in Binding rate between Time_Point 0 and Time_Point 2 is significant while the change between Time_Point 2 and Time_Point 3 is not significant. We conclude that Time_Point 2, when the monkeys had received two vaccines, had the highest Binding rate, while the last vaccine shot at Time_Point 3 did not make a difference to the Binding rate.  

## Add drugs as a covariate  

Next we add drugs as a covariate to see if it has effects on Binding. We use two indicator variables: `D2` and `D3`, where  

$$D2=
\begin{cases}
1 & \text{if Drug = 2} \\
0 & \text{otherwise}
\end{cases}
$$

$$D3=
\begin{cases}
1 & \text{if Drug = 3} \\
0 & \text{otherwise}
\end{cases}
$$

Assuming that the random effects are the same for each drug, our full model is: $$Y_{ij}= \beta_1(-2S4 + S4 Time_{ij}) + \beta_2(S4 + S5) + \beta_3(2S4 + S5 Time_{ij}) +$$ 
$$\beta_4 D2 (-2S4 + S4 Time_{ij}) + \beta_5 D2 (S4 + S5) + \beta_6 D2(2S4 + S5 Time_{ij}) +$$ 
$$\beta_7 D3(-2S4 + S4 Time_{ij}) + \beta_8 D3(S4 + S5) + \beta_9 D3 (2S4 + S5 Time_{ij}) + b_{0i} + b{1i} Time_{ij} + e_{ij}$$  

where $$\mathbf{b}_i \sim N \left( 0, \mathbf{D} = 
\begin{pmatrix}
D_{11} & D_{12} \\
 & D_{22}
 \end{pmatrix} \right)$$ and $$\mathbf{e}_i \sim N(0, \sigma^2I)$$  
 
```{r}
dataLDA1$D2 <- as.numeric(Data3$Drug == 2)
dataLDA1$D3 <- as.numeric(Data3$Drug == 3)

v1 <- dataLDA1$S4 * -2  + dataLDA1$S4 * dataLDA1$time
v2 <- dataLDA1$S4 + dataLDA1$S5
v3 <- dataLDA1$S4 * 2 + dataLDA1$S5 * dataLDA1$time
v4 <- dataLDA1$S4 * -2 * dataLDA1$D2 + dataLDA1$S4 * dataLDA1$time * dataLDA1$D2
v5 <- dataLDA1$S4 * dataLDA1$D2 + dataLDA1$S5 * dataLDA1$D2
v6 <- dataLDA1$S4 * 2 * dataLDA1$D2 + dataLDA1$S5 * dataLDA1$time * dataLDA1$D2
v7 <- dataLDA1$S4 * -2 * dataLDA1$D3 + dataLDA1$S4 * dataLDA1$time * dataLDA1$D3
v8 <- dataLDA1$S4 * dataLDA1$D3 + dataLDA1$S5 * dataLDA1$D3
v9 <- dataLDA1$S4 * 2 * dataLDA1$D3 + dataLDA1$S5 * dataLDA1$time * dataLDA1$D3

meanform3 <- binding ~ -1 + v1 + v2 + v3 + v4 + v5 + v6 + v7 + v8 + v9

fit.a3 <- lme(fixed = meanform3, random = ~ time | id, data = dataLDA1, weights = varIdent(form = ~ 1 | time), correlation = corAR1(form = ~1 | id))

summary(fit.a3)
```

Now we want to find whether the drugs have any effects. To see whether Drug 1 and Drug 2 have any difference, we want to perform a hypothesis test on $H_0: \beta_4 = \beta_5 = \beta_6 = 0$, thus we can do the test $$\mathbf{L_3 \beta}=0$$ where $$\mathbf{L_3}=
\begin{bmatrix}
0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 
\end{bmatrix}$$ and $\mathbf{\beta}=(\beta_1, \beta_2, \beta_3, \beta_4, \beta_5, \beta_6, \beta_7, \beta_8, \beta_9)^T$  

To see whether Drug 1 and Drug 3 have any difference, we want to perform a hypothesis test on $H_0: \beta_7 = \beta_8 = \beta_9 = 0$, thus we can do the test $$\mathbf{L_4 \beta}=0$$ where $$\mathbf{L_4}=
\begin{bmatrix}
0 & 0 & 0 & 0 & 0 & 0 & 1 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 0 \\
0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 
\end{bmatrix}$$ and $\mathbf{\beta}=(\beta_1, \beta_2, \beta_3, \beta_4, \beta_5, \beta_6, \beta_7, \beta_8, \beta_9)^T$  

To see whether Drug 2 and Drug 3 have any difference, we want to perform a hypothesis test on $H_0: \beta_4 = \beta_7, \beta_5 = \beta_8, \beta_6 = \beta_9$, thus we can do the test $$\mathbf{L_5 \beta}=0$$ where $$\mathbf{L_5}=
\begin{bmatrix}
0 & 0 & 0 & 1 & 0 & 0 & -1 & 0 & 0 \\
0 & 0 & 0 & 0 & 1 & 0 & 0 & -1 & 0 \\
0 & 0 & 0 & 0 & 0 & 1 & 0 & 0 & 1 
\end{bmatrix}$$ and $\mathbf{\beta}=(\beta_1, \beta_2, \beta_3, \beta_4, \beta_5, \beta_6, \beta_7, \beta_8, \beta_9)^T$  

We found that, as shown in Table \@ref(tab:L3-test), Drug 1 and Drug 2 do not have significantly different effects on Binding rates. As shown in Table \@ref(tab:L4-test), Drug 1 and Drug 3 do not have significantly different effects on Binding rates. Also, as shown in Table \@ref(tab:L5-test), Drug 2 and Drug 3 do not have significantly different effects on Binding rates. In other words, drug groups do not have signifant effects on our longitudinal model. Thus we will retain `fit.a2` as our best model.  

```{r}
betahat <- fit.a3$coefficients$fixed
V.robust <- vcovCR(fit.a3, type = "CR0")
df <- nrow(dataLDA1) - length(betahat)
```

```{r, L3-test}
# F-test
L3 <- matrix(c(0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0), byrow = TRUE, nrow = 3)
varmat3 <- L3 %*% V.robust %*% t(L3)
c3 <- nrow(L3)
est_L3 <- L3 %*% betahat
Fstat_L3 <- c( t(est_L3) %*% solve(varmat3) %*% est_L3 ) / c3
p.value <- pf(q = Fstat_L3, df1 = c3, df2 = df, lower.tail = FALSE)
knitr::kable(data.frame(Fstat = Fstat_L3, p_value = p.value), caption = "Test whether drug 1 = drug 2")
```

```{r, L4-test}
# F-test
L4 <- matrix(c(0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 1), byrow = TRUE, nrow = 3)
varmat4 <- L4 %*% V.robust %*% t(L4)
c4 <- nrow(L4)
est_L4 <- L4 %*% betahat
Fstat_L4 <- c( t(est_L4) %*% solve(varmat4) %*% est_L4 ) / c4
p.value <- pf(q = Fstat_L4, df1 = c4, df2 = df, lower.tail = FALSE)
knitr::kable(data.frame(Fstat = Fstat_L4, p_value = p.value), caption = "Test whether drug 1 = drug 3")
```

```{r, L5-test}
# F-test
L5 <- matrix(c(0, 0, 0, 1, 0, 0, -1, 0, 0, 0, 0, 0, 0, 1, 0, 0, -1, 0, 0, 0, 0, 0, 0, 1, 0, 0, -1), byrow = TRUE, nrow = 3)
varmat5 <- L5 %*% V.robust %*% t(L5)
c5 <- nrow(L5)
est_L5 <- L5 %*% betahat
Fstat_L5 <- c( t(est_L5) %*% solve(varmat5) %*% est_L5 ) / c5
p.value <- pf(q = Fstat_L5, df1 = c5, df2 = df, lower.tail = FALSE)
knitr::kable(data.frame(Fstat = Fstat_L5, p_value = p.value), caption = "Test whether drug 2 = drug 3")
```

